---
title: "Tidy US election and census data"
author: "Erik Ländström"
date: '2020-01-29'
output: 
  html_document:
    code_folding: show
    toc: yes
    toc_float: yes
    toc_depth: 4
    keep_md: yes
    theme: cerulean
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is my project for testing machine learning techniques in order to predict US election outcomes from census data.

This markdown is mainly for tidying data. 

# Load libraries

```{r libraries, message = FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
```

## Load functions

```{r functions}
source("functions/table_styling.R")
```


# Read US census data

Read US census data from 2015 (data from [www.kaggle.com](https://www.kaggle.com/muonneutrino/us-census-demographic-data)).

```{r read census data}
census_data <- read_csv("data/acs2015_county_data.csv")

# Print first rows
head(census_data) %>% 
  table_styling()

# Get a glimpse of the data
dim(census_data)
glimpse(census_data)
```

The census data has `3220` observations and `37` variables.

Read 2016 election results (data taken from [GitHub user tonmcg](https://github.com/tonmcg/US_County_Level_Election_Results_08-16/blob/master/2016_US_County_Level_Presidential_Results.csv)).

```{r read election data}
election_data <- read_csv("data/US_2016_election_results.txt")

# Print first rows
head(election_data)

# Get a glimpse of the data
dim(election_data)
glimpse(election_data)
```

The census data has `3141` observations and `11` variables, this is ``r 3220 - 3141``
fewer observations than in the census data. Let's investigate!

# Tidy data

## Prepare data sets for join

Check how many states are included in each data set.

```{r pre join 1}
# Census data
census_data %>% 
  group_by(State) %>% 
  summarise(n = n())

# Election data
election_data %>% 
  group_by(state_abbr) %>% 
  summarise(n = n())
```

The census data includes Puerto Rico, who can't vote in elections. So I will 
remove these observations from the `census_data`.

```{r puerto rico}
census_data_voting <- census_data %>% 
  filter(State != "Puerto Rico")

# Check dimensions
dim(census_data_voting)
```

There's still one more observation in `census_data_voting`. Check it out with
an `anti_join`.

```{r antijoin}
# First change name of fips code column
# I will also remove the first column X1
election_data_trimmed <- election_data %>% 
  select(-X1) %>% 
  rename(CensusId = combined_fips)

# Use anti_join to join the datasets
anti_join(election_data_trimmed, census_data_voting, by = "CensusId")
anti_join(census_data_voting, election_data_trimmed, by = "CensusId")
```

There are 5 counties that are not shared between the two data sets.

Use a `full_join` to merge the two data sets.

```{r full_join}
data_joined <- full_join(census_data_voting, election_data_trimmed, by = "CensusId")

# Check stats
dim(data_joined)
glimpse(data_joined)
```

We now have a dataset with `3144` observations and `46` variables. However,
this is two more than in the census data so there might be some missing values.

Check for missing values and remove rows including `na`s.

```{r missing values}
data_joined %>% 
  filter(is.na(total_votes))

data_joined %>% 
  filter(is.na(TotalPop))

# Remove observations with NAs
data_trimmed <- data_joined %>% 
  filter(!is.na(TotalPop)) %>% 
  filter(!is.na(total_votes))

# Check dimensions
dim(data_trimmed)
```

There are five observations that are specific to one dataset and which have been
removed. However, the county in South Dakota seems to be the same with similar 
name and different `CensusId`.

Let's see if we can join them!

```{r south dakota}
# Filter out the SD county in census
sd_census <- census_data_voting %>% 
  filter(County == "Oglala Lakota")

# Append voting data to Oglala Lakota
sd_votes <- cbind(sd_census, election_data_trimmed %>% 
        filter(county_name == "Oglala County") %>% 
          select(-CensusId))

# Append observation to the dataset
data <- rbind(data_trimmed, sd_votes)
dim(data)
```

I have no idea why there is one more Hawaii county in the census data. The data
from Alaska seems different between the two data sets.

## Alaska!

```{r alaska 1} 
# Filter for Alaska
data %>% 
  filter(State == "Alaska")
```

The voting data is the same for all counties in Alaska, while the census data
is different. I either must remove all Alaska counties or count Alaska as a
state. Since much of the data is relative or an average it would be difficult
to know if all the data is correct, so I will simply remove them.

```{r alaska 2}
# Remove Alaska observations
tidy_data <- data %>% 
  filter(State != "Alaska")
dim(tidy_data)
```

Our final dataset includes `50` states including District of Colombia and
`3112` observations and `46` columns. 

# Create binary column for who one each country

For later machine learning purposes.

```{r won}
tidy_data %>% 
  mutate(won = if_else(votes_gop > votes_dem, "Trump", "Clinton"))
```

# Save data

```{r save data, eval = FALSE}
write_tsv(tidy_data, "data/US_2016_election_and_2015_census_data.tsv")
```





















